---
title: 'Class 12: DESeq2'
author: "Jessica Diaz-Vigil"
date: "2023-05-12"
output: html_document
---

# Bioconductor and DESeq2 Setup

```{r}
#install.packages("BiocManager")
#BiocManager::install()
```

```{r}
#BiocManager::install("DESeq2")
```

```{r}
#library(BiocManager)
#library(DESeq2)
```

# Import `countData` and `colData`

```{r}
counts <- read.csv("airway_scaledcounts.csv", row.names=1)
metadata <-  read.csv("airway_metadata.csv")
```

```{r}
head(counts)
```

```{r}
head(metadata)
```

-   **Q1.** How many genes are in this dataset? 

    ```{r}
    dim(counts)
    ```

    There are 38694 genes in the dataset

-   **Q2.** How many \'control\' cell lines do we have?

    ```{r}
    control_cell_lines <- table(metadata$dex)["control"]
    control_cell_lines
    ```

    There are 4 "control" cell lines.

# Toy Differential Gene Expression

```{r}
control <- metadata[metadata[,"dex"]=="control",]
control.counts <- counts[ ,control$id]
control.mean <- rowSums( control.counts )/4 
head(control.mean)
```

```{r}
#library(dplyr)
control <- metadata %>% filter(dex=="control")
control.counts <- counts %>% select(control$id) 
control.mean <- rowSums(control.counts)/4
head(control.mean)
```

-   **Q3.** How would you make the above code in either approach more robust?

    If you had more samples, you would need to change the code for `control.mean` to being divided by total samples instead of the 4 we have now.

    ```{r}
    # control <- metadata %>% filter(dex=="control")
    # control.counts <- counts %>% select(control$id) 
    # control.mean <- rowSums(control.counts)/ TOTAL NUM
    # head(control.mean)
    ```

-   **Q4.** Follow the same procedure for the `treated` samples (i.e. calculate the mean per gene across drug treated samples and assign to a labeled vector called `treated.mean`)

    ```{r}
    treated <- metadata[metadata[,"dex"]=="treated",]
    treated.mean <- rowSums( counts[ ,treated$id] )/4 
    names(treated.mean) <- counts$ensgene
    ```

    ```{r}
    head(control.mean)
    head(treated.mean)
    ```

```{r}
meancounts <- data.frame(control.mean, treated.mean)
head(meancounts)
```

```{r}
sum.control.mean <- colSums(meancounts)["control.mean"]
sum.treated.mean <- colSums(meancounts)["treated.mean"]
sum.control.mean
sum.treated.mean
```

-   **Q5 (a).** Create a scatter plot showing the mean of the treated samples against the mean of the control samples. Your plot should look something like the following.

    ```{r}
    plot(x=control.mean, y=treated.mean, xlab= "control", ylab= "treated")
    ```

    ```{r}
    # library(ggplot2)
    ggplot(meancounts, aes(x=control.mean, y=treated.mean)) +
      geom_point()
    ```

-   **Q6.** Try plotting both axes on a log scale. What is the argument to **plot()** that allows you to do this?

    ```{r}
    ggplot(meancounts, aes(x=control.mean, y=treated.mean)) +
      geom_point() + scale_x_log10() + scale_y_log10()
    ```

```{r}
meancounts$log2fc <- log2(meancounts[,"treated.mean"]/meancounts[,"control.mean"])
head(meancounts)
```

```{r}
zero.vals <- which(meancounts[,1:2]==0, arr.ind=TRUE)
to.rm <- unique(zero.vals[,1])
mycounts <- meancounts[-to.rm,]
head(mycounts)
```

-   **Q7.** What is the purpose of the `arr.ind` argument in the **which()** function call above? Why would we then take the first column of the output and need to call the **unique()** function?

    The `arr.ind = TRUE` argument will clause `which()` to return both the row and column indices where there are true values. Were calling `unique()` since we do not want to call the same two twice just in case there are entries in both samples
