---
title: 'Class 12: DESeq2'
author: "Jessica Diaz-Vigil"
date: "2023-05-12"
output: pdf_document
---

# Bioconductor and DESeq2 Setup

```{r}
#install.packages("BiocManager")
#BiocManager::install()
```

```{r}
#BiocManager::install("DESeq2")
```

```{r}
library(BiocManager)
library(DESeq2)
```

# Import `countData` and `colData`

```{r}
counts <- read.csv("airway_scaledcounts.csv", row.names=1)
metadata <-  read.csv("airway_metadata.csv")
```

```{r}
head(counts)
```

```{r}
head(metadata)
```

-   **Q1.** How many genes are in this dataset? 

    ```{r}
    dim(counts)
    ```

    There are 38694 genes in the dataset

-   **Q2.** How many 'control' cell lines do we have?

    ```{r}
    control_cell_lines <- table(metadata$dex)["control"]
    control_cell_lines
    ```

    There are 4 "control" cell lines.

# Toy Differential Gene Expression

```{r}
control <- metadata[metadata[,"dex"]=="control",]
control.counts <- counts[ ,control$id]
control.mean <- rowSums( control.counts )/4 
head(control.mean)
```

```{r}
library(dplyr)
control <- metadata %>% filter(dex=="control")
control.counts <- counts %>% select(control$id) 
control.mean <- rowSums(control.counts)/4
head(control.mean)
```

-   **Q3.** How would you make the above code in either approach more robust?

    If you had more samples, you would need to change the code for `control.mean` to being divided by total samples instead of the 4 we have now.

    ```{r}
    # control <- metadata %>% filter(dex=="control")
    # control.counts <- counts %>% select(control$id) 
    # control.mean <- rowSums(control.counts)/ TOTAL NUM
    # head(control.mean)
    ```

-   **Q4.** Follow the same procedure for the `treated` samples (i.e. calculate the mean per gene across drug treated samples and assign to a labeled vector called `treated.mean`)

    ```{r}
    treated <- metadata[metadata[,"dex"]=="treated",]
    treated.mean <- rowSums( counts[ ,treated$id] )/4 
    names(treated.mean) <- counts$ensgene
    ```

    ```{r}
    head(control.mean)
    head(treated.mean)
    ```

```{r}
meancounts <- data.frame(control.mean, treated.mean)
head(meancounts)
```

```{r}
sum.control.mean <- colSums(meancounts)["control.mean"]
sum.treated.mean <- colSums(meancounts)["treated.mean"]
sum.control.mean
sum.treated.mean
```

-   **Q5 (a).** Create a scatter plot showing the mean of the treated samples against the mean of the control samples. Your plot should look something like the following.

    ```{r}
    plot(x=control.mean, y=treated.mean, xlab= "control", ylab= "treated")
    ```

    ```{r}
    library(ggplot2)
    ggplot(meancounts, aes(x=control.mean, y=treated.mean)) +
      geom_point()
    ```

-   **Q6.** Try plotting both axes on a log scale. What is the argument to **plot()** that allows you to do this?

    ```{r}
    library(ggplot2)
    ggplot(meancounts, aes(x=control.mean, y=treated.mean)) +
      geom_point() + scale_x_log10() + scale_y_log10()
    ```

```{r}
meancounts$log2fc <- log2(meancounts[,"treated.mean"]/meancounts[,"control.mean"])
head(meancounts)
```

```{r}
zero.vals <- which(meancounts[,1:2]==0, arr.ind=TRUE)
to.rm <- unique(zero.vals[,1])
mycounts <- meancounts[-to.rm,]
head(mycounts)
```

-   **Q7.** What is the purpose of the `arr.ind` argument in the **which()** function call above? Why would we then take the first column of the output and need to call the **unique()** function?

    The `arr.ind = TRUE` argument will clause `which()` to return both the row and column indices where there are true values. Were calling `unique()` since we do not want to call the same two twice just in case there are entries in both samples

```{r}
mycounts$log2fc <- log2(mycounts$treated.mean/ 
                        mycounts$control.mean)
up.ind <- mycounts$log2fc > 2
down.ind <- mycounts$log2fc < (-2)
```

-   **Q8.** Using the `up.ind` vector above can you determine how many up regulated genes we have at the greater than 2 fc level? 

    ```{r}
    sum(up.ind)
    ```

-   **Q9.** Using the `down.ind` vector above can you determine how many down regulated genes we have at the greater than 2 fc level? 

    ```{r}
    sum(down.ind)
    ```

-   **Q10.** Do you trust these results? Why or why not?

    I do not trust these results fully since we don't know if the changes are significant.

# DESeq2 Analysis

```{r}
library(DESeq2)
citation("DESeq2")
```

```{r}
dds <- DESeqDataSetFromMatrix(countData=counts, 
                              colData=metadata, 
                              design=~dex)
dds
```

```{r}
dds <- DESeq(dds)
```

```{r}
res <- results(dds)
res
```

```{r}
summary(res)
```

```{r}
res05 <- results(dds, alpha=0.05)
summary(res05)
```

# Adding Annotation Data

```{r}
#BiocManager::install("AnnotationDbi")
#BiocManager::install("org.Hs.eg.db") 
```

```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")
```

```{r}
columns(org.Hs.eg.db)
```

```{r}
res$symbol <- mapIds(org.Hs.eg.db,
                     keys=row.names(res),
                     keytype="ENSEMBL",
                     column="SYMBOL",
                     multiVals="first")
```

```{r}
head(res)
```

-   **Q11.** Run the **mapIds()** function two more times to add the Entrez ID and UniProt accession and GENENAME as new columns called `res$entrez`, `res$uniprot` and`res$genename`.

    ```{r}
    res$entrez <- mapIds(org.Hs.eg.db,
                         keys=row.names(res),
                         column="ENTREZID",
                         keytype="ENSEMBL",
                         multiVals="first")

    res$uniprot <- mapIds(org.Hs.eg.db,
                         keys=row.names(res),
                         column="UNIPROT",
                         keytype="ENSEMBL",
                         multiVals="first")

    res$genename <- mapIds(org.Hs.eg.db,
                         keys=row.names(res),
                         column="GENENAME",
                         keytype="ENSEMBL",
                         multiVals="first")

    head(res)
    ```

```{r}
ord <- order( res$padj )
#View(res[ord,])
head(res[ord,])
```

```{r}
write.csv(res[ord,], "deseq_results.csv")
```

# Data Visualization

```{r}
plot( res$log2FoldChange,  -log(res$padj), 
      xlab="Log2(FoldChange)",
      ylab="-Log(P-value)")
```

```{r}
plot( res$log2FoldChange,  log(res$padj), 
      xlab="Log2(FoldChange)",
      ylab="-Log(P-value)")
```

```{r}
plot( res$log2FoldChange,  -log(res$padj), 
 ylab="-Log(P-value)", xlab="Log2(FoldChange)")

# Add some cut-off lines
abline(v=c(-2,2), col="darkgray", lty=2)
abline(h=-log(0.05), col="darkgray", lty=2)
```

```{r}
mycols <- rep("gray", nrow(res))
mycols[ abs(res$log2FoldChange) > 2 ]  <- "red" 

inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2 )
mycols[ inds ] <- "blue"

plot( res$log2FoldChange,  -log(res$padj), 
 col=mycols, ylab="-Log(P-value)", xlab="Log2(FoldChange)" )

abline(v=c(-2,2), col="gray", lty=2)
abline(h=-log(0.1), col="gray", lty=2)
```

```{r}
#BiocManager::install("EnhancedVolcano")
```

```{r}
library(ggrepel)
library(EnhancedVolcano)
```

```{r}
x <- as.data.frame(res)

EnhancedVolcano(x,
    lab = x$symbol,
    x = 'log2FoldChange',
    y = 'pvalue')
```

# Pathway Analysis

```{r}
#BiocManager::install( c("pathview", "gage", "gageData") )
```

```{r}
library(pathview)
library(gage)
library(gageData)
data(kegg.sets.hs)
head(kegg.sets.hs, 2)
```

```{r}
foldchanges = res$log2FoldChange
names(foldchanges) = res$entrez
head(foldchanges)
```

```{r}
keggres = gage(foldchanges, gsets=kegg.sets.hs, same.dir=TRUE)
```

```{r}
attributes(keggres)
```

```{r}
head(keggres$less, 3)
```

```{r}
pathview(gene.data=foldchanges, pathway.id="hsa05310")
```

```{r}
pathview(gene.data=foldchanges, pathway.id="hsa05310", kegg.native=FALSE)
```

-   **Q12**. Can you do the same procedure as above to plot the pathview figures for the top 2 down-reguled pathways?
