---
title: "Class 05: Data Visualization with GGPLOT"
author: "Jessica Diaz-Vigil"
date: 2023-04-19
format: pdf
---

# GGPLOT

We are going to start by generating the plot of class 04. This code is plotting the **cars** `dataset`.

```{r}
plot(cars)
```

First, we need to install the package. We do this by using the `install.packages` command.

```{r}
# install.packages('ggplot2')
```

After that, we need to load the package.

```{r}
library(ggplot2)
```

We are going to build the plot of the cars `dataframe` by using ggplot2.

```{r}
ggplot(data = cars)
```

```{r}
ggplot(data = cars) + aes(x=speed, y=dist)

ggplot(data = cars) + 
  aes(x=speed, y=dist) + geom_point()

ggplot(data = cars) + 
  aes(x=speed, y=dist) + geom_point() +
  geom_smooth(method = 'lm')

ggplot(data = cars) + 
  aes(x=speed, y=dist) + geom_point() +
  geom_smooth(method = 'lm') + 
  labs( title = "Distance vs. Speed", 
        subtitle = 'A nice subtitle!', 
        caption = 'BIMM143', 
        x = "Speed (MPH)", 
        y = 'Distance (ft)')
ggplot(data = cars) + 
  aes(x=speed, y=dist) + geom_point() +
  geom_smooth(method = 'lm') + 
  labs( title = "Distance vs. Speed", 
        subtitle = 'A nice subtitle!', 
        caption = 'BIMM143', 
        x = "Speed (MPH)", 
        y = 'Distance (ft)') + theme_bw()
```

# PLOTTING EXPRESSION DATA

Loading the data from the URL

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"

genes <- read.delim(url)
head(genes)
```

Initial ggplot

```{r}
ggplot(data = genes) + 
  aes(x = Condition1, y = Condition2) + 
  geom_point()

nrow(genes)
ncol(genes)
colnames(genes)
table(genes['State'])
round( table(genes$State)/nrow(genes) * 100, 2 )
```

Storing the plot

```{r}
p <- ggplot(data = genes) + 
  aes(x = Condition1, y = Condition2, 
      col = State) + geom_point() 
```

Adding color to the plot

```{r}
ggplot(data = genes) + 
  aes(x = Condition1, y = Condition2, col = State) + 
  geom_point() 
```

Changing colors (pink!!)

```{r}
p1 <- p + scale_colour_manual( values=c("hotpink","deeppink","lightpink") )
p1
```

Adding Labels

```{r}
p1 + labs(title = "Differential Gene Espression", 
          x = "Control (no drug)", 
          y = "Drug Treatment", 
          caption = "BIMM 143, Class 05")
```

# LAB QUESTIONS

**Q1.** For which phases is data visualization important in our scientific workflows?

All of the above: Communication of results, Exploratory data analysis, Detection of outliers

**Q2.** True or False? The ggplot2 package comes already installed with R?

False

**Q3.** Which plot types are typically NOT used to compare distributions of numeric variables?

Network graphs

**Q4.** Which statement about data visualization with ggplot2 is incorrect? 

ggplot2 is the only way to create plots in R

**Q5.** Which geometric layer should be used to create scatter plots in ggplot2?

geom_point()

**Q6.** Use the `nrow()` function to find out how many genes are in this `dataset`. What is your answer? 

5196

**Q7.** Use the `colnames()` function and the `ncol()` function on the `genes`data frame to find out what the column names are (we will need these later) and how many columns there are. How many columns did you find?

4

**Q8.** Use the `table()` function on the `State` column of this data.frame to find out how many 'up' regulated genes there are. What is your answer?

127

**Q9.** Using your values above and 2 significant figures. What fraction of total genes is up-regulated in this `dataset`?

2.44
