---
title: "Class 18 Mini Project"
author: "Jessica Diaz-Vigil"
date: "2023-06-02"
output: pdf_document
---

# Exploring a Cancer Sequencing Data Portal

Skin Cutaneous Melanoma

### Discussion #1

**Q1.** How many cancer samples are included in the dataset?

448

**Q2.** Which is the most mutated gene?

TTN

**Q3.** Which is the most common treatment undergone by patients?

Radiation 1

# Downloading Cancer Sequencing Data

```{r}
library(maftools)
skcm = read.maf('skcm_tcga_pan_can_atlas_2018/data_mutations.txt')

mm_skcm = trinucleotideMatrix(maf = skcm, prefix = 'chr', add = TRUE, ref_genome = "BSgenome.Hsapiens.UCSC.hg19")

mm_skcm = t(mm_skcm$nmf_matrix)
```

```{r}
library(MutationalPatterns)
set.seed(11111)
```

```{r}
samples_to_plot = sample(1:ncol(mm_skcm),4)
plot_96_profile(mm_skcm[,samples_to_plot], condensed = T)
```

```{r}
mutations_in_samples = colSums(mm_skcm)
mutations_in_samples = sort(mutations_in_samples, decreasing = T)
samples_to_plot = names(mutations_in_samples)[1:4]
plot_96_profile(mm_skcm[,samples_to_plot], condensed = T)
```

```{r}
mutations_in_samples = sort(mutations_in_samples, decreasing = F)
samples_to_plot = names(mutations_in_samples)[1:4]
plot_96_profile(mm_skcm[,samples_to_plot], condensed = T)
```

```{r}
relative_mutational_profile = apply(mm_skcm, 2, prop.table)
```

```{r}
average_mutational_profile = rowMeans(relative_mutational_profile)
average_mutational_profile = data.frame(average_mutational_profile)
plot_96_profile(average_mutational_profile, condensed = T)
```

# COSMIC Reference Mutational Signatures

# Assigning Reference Mutational Signatures

```{r}
cosmic_signatures = get_known_signatures(source = 'COSMIC_v3.2')

fit_res = fit_to_signatures(mm_skcm, cosmic_signatures)

contributions = fit_res$contribution
top_contributing_signatures_abs = rowMeans(contributions)

top_contributing_signatures_abs = sort(top_contributing_signatures_abs, decreasing = T)[1:4]

top_contributing_signatures_abs
```

```{r}
relative_contributions = apply(contributions,2,prop.table)

top_contributing_signatures_rel = rowMeans(relative_contributions)

top_contributing_signatures_rel = sort(top_contributing_signatures_rel, decreasing = T)[1:4]

top_contributing_signatures_rel
```

```{r}
fit_res_strict = fit_to_signatures_strict(mm_skcm, cosmic_signatures)

fit_res_strict = fit_res_strict$fit_res

contributions_strict = fit_res_strict$contribution
```

# Visualizing Mutational Signature Assignment Results

```{r}
set.seed(11111)
samples_to_plot = sample(1:ncol(mm_skcm),4)

plot_contribution(contributions[,samples_to_plot], mode = "absolute")
```

```{r}
plot_contribution(contributions[,samples_to_plot], mode = "relative")
```

```{r}
plot_contribution_heatmap(contributions, cluster_samples = F)
```

```{r}
plot_contribution(contributions_strict[,samples_to_plot], mode = "absolute")
```

```{r}
plot_contribution(contributions_strict[,samples_to_plot], mode = "relative")
```

```{r}
plot_contribution_heatmap(contributions_strict, cluster_samples = F)
```

```{r}
set.seed(11111)
samples_to_plot = sample(1:ncol(mm_skcm),4)

plot_original_vs_reconstructed(mm_skcm[,samples_to_plot],
                               fit_res$reconstructed[,samples_to_plot], y_intercept = 0.90)
```

```{r}
plot_original_vs_reconstructed(mm_skcm[,samples_to_plot],
                               fit_res_strict$reconstructed[,samples_to_plot], 
y_intercept = 0.90)
```

### Discussion #2

**Q4.** Which is the etiology of the top absolute contributing signature for liver cancer?

Aristolochic Acid Exposure

**Q5.** Which is the most prominent mutational context for the top contributing signature in skin cancer?

C \> T

**Q6.** The etiology of the top contributing signature for lung cancer corresponds to an endogenous cellular mechanism.

FALSE

**Q7.** SBS4 is one of the most common signatures found in lung cancer and is associated with tobacco smoking.

TRUE

**Q8.** SBS7d is one of the most common signatures in skin cancer and is associated with UV light exposure and high numbers of C\>T mutations. 

FALSE
