---
title: 'Class 17: Mini Project'
author: "Jessica Diaz-Vigil"
date: "2023-05-31"
output: pdf_document
---

# Getting Started

```{r}
vax <- read.csv("covid19vaccinesbyzipcode_test.csv")
head(vax)
```

**Q1.** What column details the total number of people fully vaccinated?

```{r}
colnames(vax)
```

The column that represents the total number of people fully vaccinated it "persons_fully_vaccinated"

**Q2.** What column details the Zip code tabulation area? 

```{r}
colnames(vax)
```

The column that represents the total number of people fully vaccinated it "zip_code_tabulation_area"

**Q3.** What is the earliest date in this dataset?

```{r}
table(vax$as_of_date)
```

The earliest data in this dataset is 2021-05-05

**Q4.** What is the latest date in this dataset?

```{r}
table(vax$as_of_date)
```

The latest date in this data set is 2023-05-30

```{r}
skimr::skim_without_charts(vax)
```

**Q5.** How many numeric columns are in this dataset?

By looking at the data summary using skim, there are 14 numeric columns in the dataset

**Q6.** Note that there are "missing values" in the dataset. How many `NA` values there in the `persons_fully_vaccinated`column?

```{r}
total_na_full_vac <- sum(is.na(vax$persons_fully_vaccinated))
total_na_full_vac
```

There are 17711 NA values in the "persons_fully_vaccinated" column

**Q7.** What percent of `persons_fully_vaccinated` values are missing (to 2 significant figures)? 

```{r}
total_full_vac <- nrow(vax)
total_full_vac
```

```{r}
total_na_full_vac / total_full_vac * 100
```

The percent of missing values is 8.03%

# Working With Dates

```{r}
library(lubridate)
```

```{r}
today()
```

```{r}
vax$as_of_date <- ymd(vax$as_of_date)
```

```{r}
days_since_first <- today() - vax$as_of_date[1]
days_since_first
```

```{r}
days_in_data <- vax$as_of_date[nrow(vax)] - vax$as_of_date[1]
days_in_data
```

**Q9.** How many days have passed since the last update of the dataset?

```{r}
days_since_first - days_in_data
```

8 days have passed since the last update

**Q10.** How many unique dates are in the dataset (i.e. how many different dates are detailed)?

```{r}
unique_dates <- unique(vax$as_of_date)
unique_dates
length(unique_dates)
```

There are 125 unique dates in the dataset

# Working With Zip Codes

```{r}
#install.packages('zipcodeR')
library(zipcodeR)
```

```{r}
geocode_zip('92037')
```

```{r}
zip_distance('92037','92109')
```

```{r}
reverse_zipcode(c('92037', "92109") )
```

# Focus On The San Diego Area

```{r}
sd <- vax[ vax$county == "San Diego" , ]
head(sd)
```

```{r}
library(dplyr)
sd <- filter(vax, county == "San Diego")
nrow(sd)
```

```{r}
sd.10 <- filter(vax, county == "San Diego" &
                age5_plus_population > 10000)
```

**Q11.** How many distinct zip codes are listed for San Diego County?

```{r}
unique_zip_sd <- unique(sd$zip_code_tabulation_area)
length(unique_zip_sd)
```

There are 107 unique zip codes in San Diego

Q12. What San Diego County Zip code area has the largest population in this dataset?

```{r}
which.max(sd$tot_population)
sd$zip_code_tabulation_area[87]
```

The zip code 92154 has the largest population

**Q13.** What is the overall average (with 2 decimal numbers) "Percent of Population Fully Vaccinated" value for all San Diego "County" as of "2023-05-23"?

```{r}
sd.1 <- filter(sd.10, county == "San Diego" &
                as_of_date == "2023-05-23")
head(sd.1)
```

```{r}
mean(na.omit(sd.1$percent_of_population_fully_vaccinated))
```

**Q14.** Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of "2023-05-23"?

```{r}
hist(sd.1$percent_of_population_fully_vaccinated, main = "Histogram of Vaccination Rates Across San Diego County" , xlab = "Percent of Population Fully Vaccinated on 2023-05-23")
```

# Focus On UCSD/La Jolla

```{r}
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
```

```{r}
dim(ucsd)
```

**Q15**. Using **ggplot** make a graph of the vaccination rate time course for the 92037 ZIP code area:

```{r}
library(ggplot2)
p <- ggplot(ucsd) +
  aes(x = as_of_date,
      y = percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated")
p
```

# Comparing To Similar Sized Areas

```{r}
vax.36 <- filter(vax, age5_plus_population > 36144 &
                as_of_date == "2023-05-23")
head(vax.36)
```

**Q16**. Calculate the mean *"Percent of Population Fully Vaccinated"* for ZIP code areas with a population as large as 92037 (La Jolla) *as_of_date* "2023-05-23". Add this as a straight horizontal line to your plot from above with the `geom_hline()`function?

```{r}
mean.vax.36 <- mean(vax.36$percent_of_population_fully_vaccinated)
mean.vax.36
```

```{r}
p + geom_hline(yintercept = mean.vax.36)
```

**Q17.** What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the *"Percent of Population Fully Vaccinated"*values for ZIP code areas with a population as large as 92037 (La Jolla) *as_of_date* "2023-05-23"?

```{r}
summary(vax.36$percent_of_population_fully_vaccinated)
```

The 6 number summary is above

**Q18.** Using ggplot generate a histogram of this data

```{r}
ggplot(vax.36) +
  aes(x = percent_of_population_fully_vaccinated) + 
  geom_histogram() +
  xlab("Percent Vaccinated") +
  ylab("Count") +
  xlim(c(0,1))
```

**Q19**. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?

```{r}
vax %>% filter(as_of_date == "2023-05-23") %>%  
  filter(zip_code_tabulation_area=="92040") %>%
  select(percent_of_population_fully_vaccinated)
```

This value is below the other average we found ( 0.55 compared to 0.72)

**Q20.** Finally make a time course plot of vaccination progress for all areas in the full dataset with a`age5_plus_population > 36144`

```{r}
vax.36.all <- filter(vax, age5_plus_population > 36144)

ggplot(vax.36.all) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated, 
      group=zip_code_tabulation_area) +
  geom_line(alpha=0.2, color="hotpink") +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated",
       title="Vaccination Rate Across California",
       subtitle="Only Areas with a Population above 36k are Shown") +
  geom_hline(yintercept = mean.vax.36, linetype="dashed")
```
