---
title: "Class 10: Halloween Mini Project"
author: "Jessica Diaz-Vigil"
format: pdf
---

# Importing Candy Data

```{r}
candy_file <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"
candy = read.csv(candy_file, row.names=1)
head(candy)
```

-   **Q1**. How many different candy types are in this dataset?

    ```{r}
    nrow(candy)
    ```

    There are 85 different types of candy in this dataset.

-   **Q2**. How many fruity candy types are in the dataset?

    ```{r}
    table(candy$fruity)
    ```

    There are 38 types of candy that are fruity in the dataset. We know 1 means `True` and 0 means `False`.

# What is your favorite candy?

We can look at favorite candies by looking at larger values under the `winpercent` column.

```{r}
candy["Twix", ]$winpercent
```

-   **Q3**. What is your favorite candy in the dataset and what is it's `winpercent` value?

    ```{r}
    candy["Twix", ]$winpercent
    candy["Laffy Taffy", ]$winpercent
    ```

    My favorite type of candy is Twix which has a `winpercent` value of 81.62, but since it was already looked at, my second is Laffy Taffy which has a `winpercent` of 41.39.

-   **Q4**. What is the `winpercent` value for "Kit Kat"?

    ```{r}
    candy["Kit Kat", ]$winpercent
    ```

    The `winpercent` value for Kit Kats is 76.77.

-   **Q5**. What is the `winpercent` value for "Tootsie Roll Snack Bars"?

    ```{r}
    candy["Tootsie Roll Snack Bar", ]$winpercent
    ```

    The `winpercent` value for Kit Kats is 76.77.

    ### Skim( ) Function

    ```{r}
    #install.packages("skimr")
    library("skimr")
    ```

    ```{r}
    skim(candy)
    ```

-   **Q6**. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

    The columns `n_missing` and `complete_rate` are different from the other columns since they are only ones and zeros. This is to ensure that the data was completed adequately and in our case, they all were since there is a one in every `complete_rate` row.

-   **Q7**. What do you think a zero and one represent for the `candy$chocolate` column?

    ```{r}
    candy$chocolate
    ```

    The zero in this case means that the candy does not have chocolate in it, whereas the one in this case means that candy does have chocolate in it.

-   **Q8**. Plot a histogram of `winpercent` values

    ```{r}
    hist(candy$winpercent)
    ```

-   **Q9**. Is the distribution of `winpercent` values symmetrical?

    No, they are not. In this histogram we can see that there are a lot more candies with a `winpercent` of 30-60% and there is a decrease the higher the `winpercent` values are.

-   **Q10**. Is the center of the distribution above or below 50%?

    Just by looking, we can see that the center of the distribution is below 50%.

-   **Q11**. On average is chocolate candy higher or lower ranked than fruit candy?

    ```{r}
    chocolate <-candy$winpercent[as.logical(candy$chocolate)]
    fruity <- candy$winpercent[as.logical(candy$fruity)]
    t.test(chocolate, fruity)
    ```

    ```{r}
    boxplot(chocolate, fruity)
    ```

    By looking at the means, the mean of the chocolate candies are much higher that the means of the fruity candies (60.92% compared to 44.11%).

-   **Q12**. Is this difference statistically significant?

    The difference is statistically significant since the p-value for the Welch Two Sample t-test is 2.871e-08 which is lower than 0.05.

# Overall Candy Rankings

```{r}
head(candy[order(candy$winpercent),], n=5)
```

```{r}
tail(candy[order(candy$winpercent),], n=5)
```

-   **Q13**. What are the five least liked candy types in this set?

    The least favorite candies are Nik N Lip, Boston Baked, Chiclets, Super Bubble, and Jawbusters.

-   **Q14**. What are the top 5 all time favorite candy types out of this set?

    The top favorite candy types are Snickers, Kit Kat, Twix, Reese's Miniature Cups, Reese's Peanut Butter Cups.

-   **Q15**. Make a first barplot of candy ranking based on `winpercent` values.

    ```{r}
    library(ggplot2)

    ggplot(candy) + 
      aes(winpercent, rownames(candy)) +
      geom_col() 
    ```

-   **Q16**. This is quite ugly, use the `reorder()` function to get the bars sorted by `winpercent`?

    ```{r}
    ggplot(candy) + 
      aes(winpercent, reorder(rownames(candy),winpercent)) +
      geom_col() 
    ```

    ### Adding Colors

    ```{r}
    my_cols=rep("black", nrow(candy))
    my_cols[as.logical(candy$chocolate)] = "chocolate"
    my_cols[as.logical(candy$bar)] = "brown"
    my_cols[as.logical(candy$fruity)] = "pink"
    ```

    ```{r}
    ggplot(candy) + 
      aes(winpercent, reorder(rownames(candy),winpercent)) +
      geom_col(fill=my_cols) 
    ```

-   **Q17**. What is the worst ranked chocolate candy?

    The worst ranked chocolate candy was Sixlets

-   **Q18**. What is the best ranked fruity candy?

    The worst ranked fruity candy was Nik L Nip

# Taking a look at `Pricepercent`

```{r}
#install.packages("ggrepel")
library(ggrepel)

ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

-   **Q19**. Which candy type is the highest ranked in terms of `winpercent`for the least money - i.e. offers the most bang for your buck?

    ```{r}
    ord <- order(candy$pricepercent, decreasing = TRUE)
    tail( candy[ord,c(11,12)], n=5 )
    ```

    Considering that good candies have a `winpercent` higher than 80% and candies that are cheap have a `pricepercent` lower than .3, the candy with the most bang for your buck is Reece's Miniatures. Overall, fruity candies offer more bang for your buck though (higher win values, lower price values.

-   **Q20**. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

    ```{r}
    ord <- order(candy$pricepercent, decreasing = TRUE)
    head( candy[ord,c(11,12)], n=5 )
    ```

    Considering that bad candies have a `winpercent` lower than 40% and candies that are expensive have a `pricepercent` higher than .75, the candies that are least popular and most expensive are: Nik L Nip, Ring pop, Nestle Smarties, Sugar Babies, and Pop Rocks.

# Exploring the Correlation Structure

```{r}
#install.packages("corrplot")
library(corrplot)
```

```{r}
cij <- cor(candy)
corrplot(cij)
```

-   **Q22**. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

    The two variables that are anti-correlated are fruity and chocolate.

-   **Q23**. Similarly, what two variables are most positively correlated?

    The two variables that are correlated are chocolate and bar.

# Principal Component Analysis

```{r}
pca <- prcomp(candy, scale=T)
summary(pca)
```

```{r}
plot(pca$x[,1:2])
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)
p
```

```{r}
p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
#install.packages("plotly")
library(plotly)
```

```{r}
#ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

-   Q**24**. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

    The variables that are picked up strongly by PC1 in the positive direction are fruity, hard, and pluribus meaning that these values are more correlated together than the other variables. Fruity candies tend to be hard and come in packages of many.
